Type Evolution in Programming Systems

### 2024-05-27 paper.tex
Possibly complete related work

\section{Live Modeling Languages Require Run-Time State Migration}
%Related work
%``Slogan'': editing a program is diverging from run-time schema.

@@

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help. SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes. MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards. In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

@@

@@

\paragraph{Model Driven Engineering.}
Model Driven Engineering has also studied evolution. Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our Extract Entity and Multiplicty Change examples, and would make for a good follow-on problem. However they do not address code/query rewriting.

\paragraph{Program synthesis.}
\citet{wang19} use program synthesis to rewrite SQL programs to adapt to a schema change. This process is driven by finding possible correspondences between columns in the old and new schema guided by heuristics on textual similarity. It is assumed the migration does not change datatypes or lose any information visible to the programs yet those are real needs. It is not clear how pathologically ambiguous schema changes are resolved (for example renaming equityped columns).

\paragraph{Migration by example.}
Programming by example has been used to infer schema migration programs~\cite{wang20, Alexe11}. This work has dealt with the reality that schema live in the database and so examples must be supplied separately, but we wonder why not incorporate examples into a schema definition language? Then purely textual edits that properly adapted the examples could be used to infer migrations and resolve the ambiguities from comparing pure schema.

\paragraph{Tijs} \cite{vanRozen15, vanRozen19, RuntimeConstraint, SemanticDeltas}

\paragraph{Live, MDD, etc. programming?}

% Some references from Tomas
\paragraph{collaborative programmign tools}
\cite{goldman2011real,kurniawan2015coder,replit}
\paragraph{exploratory/live review paper}
\cite{rein2018exploratory}
\paragraph{hot reload in live systems}
\cite{barenz2020essence,beckmann2021shortening}

\begin{comment}

Convex migration DSL: https://stack.convex.dev/migrating-data-with-mutations

dataframe diff https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf

Durable computing: pins runtime data frames to code versions. Doesn't handle problem of global data schema change.
https://restate.dev/blog/solving-durable-executions-immutability-problem/
https://restate.dev/blog/code-that-sleeps-for-a-month/

From Geoffrey:
1) Finding Data Compatibility Bugs with JSON Subschema Checking
https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant. It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
2) Confluent Schema Registry:
https://docs.confluent.io/platform/current/schema-registry/index.html
An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
\end{comment}

>>>>

\section{Live Modeling Languages Require Run-Time State Migration}
\label{livemodeling}
%Related work
%``Slogan'': editing a program is diverging from run-time schema.

@@

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases, which provide surprisingly little help.
SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

@@

\citet{wang19} use program synthesis to rewrite SQL programs to adapt to a schema change. This process is driven by finding possible correspondences between columns in the old and new schema guided by heuristics on textual similarity and confirmed by the existence of a synthesized equivalent program.
\begin{comment}
  I hate this paper - incredibly sophisticated techniques applied to an over-idealized problem. It is assumed the migration does not change datatypes or lose any information visible to the programs. It assumes the database has full code coverage. It ignores without explanation ambiguous schema changes (for example renaming equityped columns).
\end{comment}

@@

\emph{Local-first software}~\cite{localfirst} adds the divergence dimension to NoSQL evolution, because the code running on different replicas can evolve at different paces. \citet{Cambria} applies lenses~\cite{Foster2007} to this problem, specifically discussing the \emph{Multiplicity Change} problem.

@@

\paragraph{Model Driven Engineering.}
Model Driven Engineering must also handle evolution. Unlike the textual artifacts involved in other domains, models typically assign unique indentifiers enabling more precise differencing and mergeing~\cite{alanen2003}.
Models are often themselves modeled by a metamodel, which lifts the evolution problem to the metalevel: models must be migrated through the evolution of their metamodel, analogously to changing the grammar of a programming language.~\cite{Herrmannsdoerfer11}. \citet{Cicchetti11} study metamodel evolution in the context of divergent changes.
\citet{vermolen11} define a DSL for evolving a data model by generating SQL migrations on the backing database. They consider the evolution of OO-style subclass hierarchies which goes beyond most other work. Their running example includes both our Extract Entity and Multiplicty Change examples, and would make for a good follow-on problem. However they do not address code/query rewriting.

\paragraph{Migration by example.}
Programming by example has been used to infer schema migration programs from examples specified by the developer~\cite{wang20, Alexe11}.
This work deals with the reality that schema live in the database and so examples must be supplied separately, but we wonder why not incorporate examples into a schema definition language? Textual edits to schema in this language would also adapt the examples, which then could be used to infer an abstract migration, thus avoiding the ambiguities of textually comparing pure schema.

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (mostly solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
% A special case of live programming is \emph{hot reloading} (also called \emph{hot swapping})~\cite{barenz2020essence, } that installs code changes into a running system preserving state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in Section \ref{livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.

%\paragraph{collaborative programming tools}\cite{goldman2011real,kurniawan2015coder,replit}
